---
title: "Class 5: Data Viz with ggplot"
author: "Michael A18135877"
format: pdf
---

Today we ar eexploring the **gggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass this to the `plot()` function.

```{r}
plot(cars)
```


>Key-point: Base R is quick but not so nice looking in some folks eyes.

Let's see how we can plot this with **ggplot2**...

1st I need to install this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal. 

2nd We need to load the package with `library()` function every time we want to use it.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

- **data** (i.e a data.frame with the things you want to plot), 
- aesthitics **aes()** that map the columns of data to your plot features (i.e. aesthitics)
- geoms like **geom_point()** that sort how the plot appears

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

>Key point: For simple "canned" graphs base R is quicker but as things get more custom and elaborate then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y
Add a title
Add custom axis labels "Speed (MPH)" and "Distance (ft)"
Chnage the theme...

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(title="Silly plot of Speed vs Stopping distance") + 
  labs(x="Speed (MPH)", y="Distance (ft)")
  theme_bw()
```


## Going further

Read some expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset?

```{r}
nrow(genes)
```


> Q2. How many "up" regulated genes are there?

```{r}
table(genes$State)
```

A useful function for counting up occurances of things in a vector is the `table()` function.

Make a v1 figure

```{r}
p<-ggplot(genes) +
  aes(x= Condition1, 
      y= Condition2,
      col=State) +
  geom_point() +
  labs(title= "Expression chnages upon drug treatment", 
       x="Control (no drug)",
       y="Treatment (with drug)")
  p + scale_colour_manual(values = c("blue", "gray","red"))
```

##More PLotting
library(gapminder)

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Lets have a wee peak

```{r}
head(gapminder, 3)
```

>Q4. How many different country values are in this dataset?

```{r}
nrow(gapminder)
```

```{r}
length(table(gapminder$country))
```

>Q5. How many different continent values are in this dataset.

```{r}
unique(gapminder$continent)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color=continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color=continent, label=country) +
  geom_point() 
```

I can use the **ggrepel** package to make more sensible labels here.

```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color=continent, label=country) +
  geom_point() +
  geom_text_repel()
```

I want a separate panel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color=continent, label=country) +
  geom_point() +
  geom_text_repel() +
  facet_wrap(~continent)
```

ggplot2 offers several advantages over base R plots:

1. **Layered Grammar**: ggplot2 uses a consistent, layered approach for building plots, making it easier to add or modify elements (data, aesthetics, geoms) step by step [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

2. **Publication Quality**: ggplot2 produces visually appealing, publication-quality figures with sensible defaults, which are harder to achieve with base R without extensive tweaking [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

3. **Customization**: Customizing complex plots is more straightforward in ggplot2. Adding legends, themes, or combining multiple plots is easier and less error-prone than in base R [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk).

4. **Consistency**: The syntax and logic are consistent across different plot types, reducing the need to learn many separate functions as in base R [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

5. **Extensibility**: ggplot2 is part of a larger ecosystem, allowing integration with other packages for advanced graphics and analysis [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk).

What do you think is the most important advantage for your own work?
