---
title: "Class 6: R Functions Lab"
author: "Michael Romero (A18135877)"
format: pdf
---

All functions in R have at least 3 things:

-A **name** we pick this and use it to call the function. - Input **arguments**, there can be multiple comma separated inputs ot the function - The **body**, lines of R code that do the work of the function

Our first wee function:

```{r}
add<-function(x, y=1) {
  x+y
}
```

Lets test our function

```{r}
add(c(1,2,3), y=10)
```

```{r}
add(10,100)
```

## A second function

Let's try something more interesting. Make a sequence generation tool.

The `sample()` function could be useful here.

```{r}
sample(1:10, size= 3)
```

change this to work with the nucleotides A C G and T

```{r}
n<-c("A","C","T","G")
sample(n, size=5, replace = TRUE)
```

TUrn this snippet into a function that returns a user specified length dna sequence. Let's call it `generate_dna()`

```{r}
generate_dna<-function(len=10, fasta=FALSE) {
  n<-c("A","C","T","G")
  v<- sample(n, size=len, replace = TRUE)
  
  # Make a single element vector
  s<-paste(v, collapse="")
  
  cat("Well done you!")
  
  if(fasta) {
 return(s)    
  }else{
    return(v)
  }
}
```

```{r}
generate_dna(5)
```

```{r}
s<-generate_dna(15)
s
```

I want the option to return a single element character vector with my sequence all together like this: "GGAGTAC"

```{r}
generate_dna(10, fasta=FALSE)
```

```{r}
generate_dna(10, fasta=TRUE)
```

## A more advanced example

Make a third function that generates protein sequence of a user specified length and format. 

```{r}
generate_dna<-function(size=20, fasta=FALSE) {
  n<-c("A","C","T","G")
  v<- sample(n, size=size, replace = TRUE)
  
  # Make a single element vector
  s<-paste(v, collapse="")
  
  cat("Well done you!")
  
  if(fasta) {
 return(s)    
  }else{
    return(v)
  }
}
```

```{r}
generate_dna(6)
```

```{r}
generate_protein<-function(size=15, fasta=TRUE) {
  aa<-c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")
  seq<-sample(aa, size=size, replace=TRUE)
  
  if (fasta) {
    return(paste(seq, collapse=""))
  } else{
    return(seq)
    }
}
```


```{r}
generate_protein(5)
```

>Q. generate random protein sequences between lengths 5 and 12 amino acids.


One approach is to do this by brute force calling our function for each length 5 to 12.

Another approach is to write a `for()` loop to itterate over the input valued 5 to 12

A very useful third R specific approach is to use the `sapply()` function.




```{r}
sapply(5:12, generate_protein)
```

>**Key-Point**: Writing functions in R is doable but not the easiest thing. Starting with a working snippet of code and then using LLM tools to imporve and generalize your function code is a productive approach. 