---
title: "Class 10: Haloween mini project"
author: "Michael Romero A18135877"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our come from the 538 website and is available as a CSV file:

```{r}

candy_file <- read.csv("candy-data.txt")
```



```{r}
candy = read.csv("candy-data.txt", row.names=1)
head(candy)
```

```{r}
flextable::flextable(head(candy,10))
```


The functions dim(), nrow(), table() and sum() may be useful for answering the first 2 questions.

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
library(tidyverse)
candy |>
ncol()
```

```{r}
library(tidyverse)

candy %>%
nrow()
```

Count how many candy are fruity?

```{r}
sum(candy$fruity)
```

My favorite winpercent

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Almond Joy",]$winpercent
```

```{r}
library(dplyr)

candy |>
filter(rownames(candy)=="Almond Joy") |>
select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

##Quick overview of the dataset

```{r}

skimr::skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale the rest are 0-1 scale.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate

>Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy, aes(x=winpercent)) +
  geom_histogram()
```


>Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```
The peak is before 50 and thus is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```
The center of distribution is below 50 since the median is at 47.83% win percent. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#1. Find all chocolte candy in the dataset
#2. Find their winpercent values
#3. Calculate the mean of these values

#4-6. Do the same for fruity candy
#7. Compare mean winpercents of chocolate vs fruity
#8. Pick the highest as winner

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds,]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```
```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```
```{r}
fruit.ind <- candy$fruity==1
fruit.win <- candy[fruit.ind,]$winpercent
fruit.mean <- mean(fruit.win)
fruit.mean
```
```{r}
candy |>
  filter(chocolate==1) |>
  select(winpercent) 
```


>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
The difference is significant as p-value < 0.05.

>Q13. What are the five least liked candy types in this set?

```{r}
ord.ind <- order(candy$winpercent)
head( candy[ord.ind,], 5 )
```

```{r}
candy %>% arrange(ascending=winpercent) %>% head(5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(descending=candy$winpercent) %>% tail(5)
```

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col()
```
Add some color based on the type of candy

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

##Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps=7)

```


> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlet.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starburst.

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types are fruity, fruity, chocolate, chocolate bar, and chocolate bar. The least popular one here is the fruity candy "Nik L Nip".

##Exploring the correlation structure

Now that we've explored the dataset a little we'll see how the variables interact with each other. ... the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are strongly anti-correclated.

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are the most positively correlated.

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label=rownames(pca$x)) + 
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols)
```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs


```{r}
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) + 
  geom_col()
```
>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are strongly picked up by PC1 in the positive direction which makes sense since many fruity candy tend to be hard and contain a higher amount of them per package.

Intercative plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF :-(

```{r}
library(plotly)

#plotly(p)
```

